---
title: "influenceAnalysis_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{influenceAnalysis_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Set a CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

```

```{r}
library(influenceAnalysis)
```

# Introduction

The influenceAnalysis package provides tools for conducting influence diagnostics in linear regression models. This vignette demonstrates how to use the package to identify influential data points using various measures such as Cook's Distance, DFFITS, and Hadi's Influence Measure.

#Installation

To install the the package from GitHub:
```{r}
# Install the devtools 
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}


# Install influenceAnalysis from GitHub
devtools::install_github("HebaAleterji/Assignment2")
```


#Loading the Package and Data

First, load the influenceAnalysis package and the dataset you want to analyze. We'll use the built-in mtcars dataset for demonstration purposes.

```{r}
library(influenceAnalysis)
data(mtcars)

```

# Fitting a Linear Model

Fit a linear model that you want to analyze. For instance, let's predict miles per gallon (mpg) based on weight (wt) and horsepower (hp).

```{r}
model <- lm(mpg ~ wt + hp, data = mtcars)
summary(model)

```


# Performing Influence Diagnostics

You can now perform influence diagnostics using the `influence_diagnostics` function. This function allows you to compute various influence measures and visualize them.

```{r}
results <- influence_diagnostics(mtcars, model, "all")
```


# Plotting Influence Measures

The `plot_influence` function generates plots that help you visualize the influence of individual observations.

```{r}
plot_influence(results)
```

# Interpreting the Results
The results are stored in a list, where each element corresponds to a different influence measure (Cook's Distance, DFFITS, Hadi's Measure).

```{r}
print(results)
```


#  Working with Specific Influence Measures
If you're interested in computing a specific influence measure, the influence_diagnostics function allows you to do so directly. The available influence measures include:

- Cook's Distance: A measure of the influence of each observation on the fitted values.

- DFFITS: A measure of the change in the predicted value when a particular observation is omitted.

- Hadi's Influence Measure: A measure that considers both the distance and the leverage of the observation.

You can specify which measure to compute by passing the appropriate argument to the influence_diagnostics function. For example, to compute only Cook's Distance:

```{r}
# Compute only Cook's Distance
cooks_distances <- influence_diagnostics(mtcars, model, "cooks")

# Compute only DFFITS
dffits_values <- influence_diagnostics(mtcars, model, "dffits")

# Compute only Hadi's Influence Measure
hadi_values <- influence_diagnostics(mtcars, model, "hadi")


```

